A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-09-25 jacobian_calc_CMAME.tex
fixed typos and label syntax for table
editing and additing references

\begin{abstract} %% Text of abstract Shown is a retrospective comparative study of tangent-stiffness
matrix calculation methods including a newly developed Complex Taylor Series Expansion
(CTSE)-based ``complex-step'' method alongside established methods: forward-difference,
central-difference and automatic differentiation.  Tangent-stiffness matrix is a term from
computational mechanics which here refers to the Jacobian matrix as used in first-order
Newton-Raphson methods for the solution of non-linear algebraic systems of equations developed
to describe physical or theoretical systems studied in math, science, engineering and business.
To perform the comparative study, the above tangent-stiffness calculation methods were applied
in a massively parallel computational peridynamics code developed at Sandia National Labs,
called \emph{Peridigm}. In the comparative study, for each datum for each run, complex-step was
multiple orders of magnitude more accurate the finite-difference methods, according to a common
comparison made to the automatic-differentiation method datum. However, for the implementation
contemporary to the study, complex-step was also the slowest method among the four for each
datum for each run. A mathematical definition of complex-step and
and automatic differentiation, description of the computer implementations and justification
of the comparative study precede results for clarity. The intended audience of this paper
includes researchers, professionals and students, therefor readers are directed to
selected background sources in the text for well-known concepts while less familiar concepts
are derived in-paper with an attempt made to lay steps out plainly for those in disparate fields
or who are new to their studies.  \end{abstract}

@@

The scope of the application component of the study was limited to a single material model implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the new method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of finite-difference, central-difference, and automatic-differentiation methods needed for comparison to the new complex-step methed was developed, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters.

@@

Automatic-differentiation or 'AD' is a computerized method for computing exact derivatives based the
chain-rule from calculus. AD takes advantage of the fact that any
mathematical function executed on a computer, no matter how complicated, is a "composition of simple
operations" (add, multiply, power, transcendental and the like) each having known analytical derivatives
\cite{ref-sacado-presentation}. For reference, the AD
implementation used in the study is the "Sacado" package from the "Trilinos" library developed out of Sandia
National Labs \cite{ref-Sacado}.

@@

\subsubsection{Calculating Tangent-stiffness Matrices with Each Method}
For reference, expressions for calculating tangent stiffness matrices with CTSE, AD,
central-difference and forward-difference will be shown and explained. The expressions for AD,
central-difference and forward-difference are taken from {ADDAGIO MANUAL REF}, while the expression
for CTSE is original.

@@

Where $K_{ij}$ is the index notation representation of an element of the tangent-stiffness matrix at
row $i$, column $j$. The expression says that the partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the quotient of the partial difference of two terms
and a denomenator. The first of the 'two terms' is $F_i$ evaluated at $x$ where
the $j$'th component of $x$ has been perturbed by an very small amount in the direction of the
positive $x_j$ axis. The second of the 'two terms' is $F_I$ evaluated at an unperturbed $x$. The
denominator is the magnitude of perturbation called "probe distance" {ADDAGIO MANUAL REF}

@@

Mesh generation for the test problems run in the study was done using the software package
\emph{Cubit} developed at REFERENCE. For the test series in which peridynamic node density was
varied, journal scripts compatible with \emph{Cubit} were used. The rule used to select the
arangement of nodes within the borders designated by simulated material block's dimensions was to
manually choose a node spacing, fit the maximum whole number of nodes in a regular grid within the
block, and then check whether or not this number of nodes matched the number of nodes desired for a
particular test. The same process was used for the multicore series. An example journal script can
be found on the corresponding author's website. For future work, it may be beneficial to automate
and randomize this process for statistic reasons and for time efficiency.

@@

Since samples were not repeated, because there was not resources for it, confidence intervals were
not calculated. However, if results of iterations within the same test problem were considered
samples of the same thing, then a confidence interval could be calculated. This assumption was made
for the purpose of plotting reporting average values, but a confidence interval was not calculated.
But, a cursory review of data showed no range overlap among the methods for accuracy or speed data
from any test run of either series.

@@

\begin{table}
    \begin{tabular}{c c c c }
        \caption{Averaged Results, Each Test} \label{tab:results}
    \hline
    Cores & Jacobian Els. & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1.99E6  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{table}

@@

For the multicore tests, averaged speed results appear in \ref{fig:multi_speed}. For this series,
rather than varying the number of nonzero Jacobian matrix elements the number of cores used to
solve the problem is varied, while the number of nonzero Jacobian matrix elements remained constant.
Complex-step again was the slowest method, followed by central-difference, automatic-differentiation
and then forward-difference, taking the title of fastests method for this group of tests.
\ref{fig:multi_speed} indicates that when looking at average in-iteration speed, each of the
methods scale linearly with the number of cores used for the given conditions of this group of
tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale
well when additional processors are used. An explanation as to why automatic-differentiation and
forward-differences swapped places in speed ranking relative to the single core results was not
determined.

@@

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in \red{tcs}. The averaged
results of these accuracy measurements for the single core tests can be seen in
\ref{fig:serial_accuracy}. Averaged Frobenius norm of the element-wise difference between a given
Jacobian and the Jacobian produced by the automatic-differentiation method is plotted against number
of nonzero Jacobian matrix elements. Complex-step was revealed as the most accurate of the methods
besides automatic-differentiation, complex-step resulting in several orders of magnitude less
difference that resulted from using the central-difference method and the most inaccurate
forward-difference method. This accuracy ranking order was mantained for each test in the series. It
should be pointed out compared to the L2 norms of any of the Jacobian matrices by themselves, the
magnitude of these differences for any of the methods is relatively small. These values appear in
the experimental data which is available on the corresponding author's website. The effect of these
accuracy differences on convergence rate directly was not determined.

@@

\subsection{Thoughts on Results} Based on the experimental results, it appears that for users of
\emph{Peridigm} there is a case to be made for using automatic-differentiation for reasons of
performance and accuracy. Although, the greater speed of finite-difference for certain multicore
tests done in this study may give some users pause. While the results showed that complex-step was
clearly more accurate than central-difference and forward-difference for the parameters of the
tests, it was not determined whether or not this is a clear advantage of complex-step over
finite-difference in terms of accuracy of final results and speed of convergence.   For other
simulation software it is obviously necessary to collect the same types measurements as decribed in
this paper in order to make a design decision. For new developments it is recommended that combined
accuracy and speed of iteration mesurments be conduted along with entirely separate convergence
studies.  For reasons explained in the next subsection, a convergence study in \emph{Peridigm} performed using
the implementations of the methods used for this paper would be moot.

@@

After the bulk of the study had been conducted, gained familiarity with \emph{Peridigm} revealed
that the implementations of forward-difference, central-difference and by virtue of reusing code,
complex-step, in \emph{Peridigm} were not optimal due to inefficiencies in the implementations of
the Linear Peridynamic Solid (LPS) model. Each of these implementations included extensive looping
over noncontiguous addresses in memory and unnecessary re-calculation of constant values.  The
approach used does minimize temporary memory use, and preserve modularity of the code for non-LPS
models, but is very wasteful of cpu and memory bandwidth for use with LPS. Unpublished preliminary
results show that when calculations for the quantity 'zeta' are re-used and only perturbed and
previously pertubed functions of independent variables are recomputed throughout the course of a
Jacobian evaluation, complex-step performs at the speed of unmodified forward-difference. Because of
this it is speculated that forward-difference and central-difference could be modified in nearly the
same way to perform more quickly than AD. In fact, because forward difference involves double the
number of model evaluations than that of complex-step, forward-difference stands to gain
proportionally more efficiency. Although no formal academic paper will be likely written based on
this, a brief report and an enhanced performance implementation of forward-difference for peridynamic LPS
in \emph{Peridigm} will eventually be made available on the corresponding author's website and is
intended to be submitted for official inclusion in future releases of \emph{Peridigm}.

\subsection{Final Thoughts on Complex-Step} Complex-step was shown to be highly accurate by the
standards of this study, but as it is implemented in \emph{Peridigm} may not be any better than AD,
and it is almost certainly slower for a naive implementation. When looking to the future of
\emph{Peridigm}, Complex-step holds the advantage of relying on byte-copiable datatypes, while AD
requires a complicated serialization and deserialization in order to function as byte-copiable. This
is fine for copying in between MPI processes, but deserialization on an accelerator or GPU is very
difficult, as this would require compiling a version of the AD library being used that were
compatible with the accelerator or GPU being used. What this means is that if \emph{Peridigm} or any
simulation software were to be reformulated to take advantage of accelerators or GPUs, attractive
given their performance and power efficiency advantages over traditional CPUs, the flexibilty of
complex-step owed to its simple implementation may make it more viable fo rthat application.

>>>>

\begin{abstract} %% Text of abstract Shown is a retrospective comparative study of tangent-stiffness
matrix calculation methods including a newly developed Complex Taylor Series Expansion
(CTSE)-based ``complex-step'' method alongside established methods: forward-difference,
central-difference and automatic differentiation.  Tangent-stiffness matrix is a term from
computational mechanics which here refers to the Jacobian matrix as used in first-order
Newton-Raphson methods for the solution of non-linear algebraic systems of equations developed
to describe physical or theoretical systems studied in math, science, engineering and business.
To perform the comparative study, the above tangent-stiffness calculation methods were applied
in a massively parallel computational peridynamics code developed at Sandia National Labs,
called \emph{Peridigm}. In the comparative study, for each datum for each run, complex-step was
multiple orders of magnitude more accurate the finite-difference methods, according to a common
comparison made to the automatic-differentiation method datum. However, for the implementation
contemporary to the study, complex-step was also the slowest method among the four for each
datum for each run. A mathematical definition of complex-step and
and automatic differentiation, description of the computer implementations and justification
of the comparative study precede results for clarity. The intended audience of this paper
includes researchers, professionals and students, therefore readers are directed to
selected background sources in the text for well-known concepts while less familiar concepts
are derived in-paper with an attempt made to lay steps out plainly for those in disparate fields
or who are new to their studies.  \end{abstract}

@@

The scope of the application component of the study was limited to a single material model implemented in the computational peridynamics code, \textit{Peridigm} \cite{peridigm}. Identifying a specific application provided a practical framework for implementing the new method and solving engineering problems to generate the data needed to compare the methods. In particular, \emph{Peridigm} was chosen because it combined several helpful characteristics: the utilization of Newton's method and tangent-stiffness matrices for solving nonlinear quasi-static problems, prior inclusion of finite-difference, central-difference, and automatic-differentiation methods needed for comparison to the new complex-step method was developed, and being an agile-components code that makes use of distributed computing data structures via Trilinos \cite{trilinos} for efficient parallelization on large clusters.

@@

Automatic-differentiation or 'AD' is a computerized method for computing exact derivatives based the
chain-rule from calculus. AD takes advantage of the fact that any
mathematical function executed on a computer, no matter how complicated, is a "composition of simple
operations" (add, multiply, power, transcendental and the like) each having known analytical derivatives
\cite{ref-sacado-presentation}. For reference, the AD
implementation used in the study is part of the \emph{Sacado} package from the \emph{Trilinos} library developed out of Sandia National Labs \cite{ref-Sacado}.

@@

\subsubsection{Calculating Tangent-stiffness Matrices with Each Method}
For reference, expressions for calculating tangent stiffness matrices with CTSE, AD,
central-difference and forward-difference will be shown and explained. The expressions for AD,
central-difference and forward-difference are taken from \cite{ref-Adaggio}, while the expression
for CTSE is original.

@@

Where $K_{ij}$ is the index notation representation of an element of the tangent-stiffness matrix at
row $i$, column $j$. The expression says that the partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the quotient of the partial difference of two terms
and a denomenator. The first of the 'two terms' is $F_i$ evaluated at $x$ where
the $j$'th component of $x$ has been perturbed by an very small amount in the direction of the
positive $x_j$ axis. The second of the 'two terms' is $F_I$ evaluated at an unperturbed $x$. The
denominator is the magnitude of perturbation called "probe distance" \cite{ref-Adaggio}

@@

Mesh generation for the test problems run in the study was done using the software package
\emph{Cubit} developed at Sandia National Labs \cite{ref-Cubit}. For the test series in which peridynamic node density was
varied, journal scripts compatible with \emph{Cubit} were used. The rule used to select the
arangement of nodes within the borders designated by simulated material block's dimensions was to
manually choose a node spacing, fit the maximum whole number of nodes in a regular grid within the
block, and then check whether or not this number of nodes matched the number of nodes desired for a
particular test. The same process was used for the multicore series. An example journal script can
be found on the corresponding author's website. For future work, it may be beneficial to automate
and randomize this process for statistic reasons and for time efficiency.

@@

Since samples were not repeated, because there was not resources for it, confidence intervals were
not calculated. However, if results of iterations within the same test problem were considered
samples of the same thing, then a confidence interval could be calculated. This assumption was only made
for the purpose of reporting average values on the plots. But, a cursory review of data showed no 
range overlap among the methods for accuracy or speed data from any test run of either series.

@@

\begin{table}    
        \caption{Averaged Results, Each Test} \label{tab:results}   
    \begin{tabular}{c c c c }
    Cores & Jacobian Els. & Calc. Time (s) & Accuracy Diff. (MPa)\\ 
    \multicolumn{2}{c}{} & CS, CD, FD, AD & CS, CD, FD \\ \hline 
	1 & 1.99E6  & 3.5, 2.7, 1.7, 1.6 & 1.92E-10, 1.21E-4, .137 \\ \hline
	1 & 4.25E6  & 6.2, 4.9, 3.2, 2.9 & 2.28E-10, 9.94E-4, .148 \\ \hline
	1 & 7.79E6  & 11.2, 8.9, 5.7, 5.2 & 2.38E-10, 1.59E-4, .145\\ \hline
	1 & 1.54E7  & 26.7, 21.0, 13.4, 12.2 & 2.33E-10, 4.61E-4, .12 \\ \hline
	1 & 2.00E7  & 28.1, 22.2, 14.4, 13.1 & 2.76E-10, 1.05E-3, .145 \\ \hline
	1 & 3.14E7  & 47.6, 37.7, 24.2, 21.9 & 2.64E-10, 1.65E-3, .133 \\ \hline
	1 & 4.01E7  & 55.6, 44.1, 28.4, 25.9 & 3.03E-10, 1.92E-3, .148 \\ \hline
	1 & 8.39E7  & 138.9, 109.6, 70.2, 64.0 & 3.63E-10, 1.64E-3, .123 \\ \hline
	1 & 1.65E8  & 277.3, 218.1, 139.4, 126.5 & 3.26E-10, 2.18E-3, .128 \\ \hline
	32 & 1.67E10  & 336.1, 277.7, 200.6, 233.1 & 6.21E-10, 1.52E-2, .176 \\ \hline
	64 & 1.67E10  & 169.9, 140.7, 102.0, 119.7 & 6.20E-10, 1.50E-2, .177 \\ \hline
	96 & 1.67E10  & 114.7, 95.0, 69.1, 79.7 & 6.18E-10, 1.50E-2, .177 \\ \hline
	128 & 1.67E10  & 86.4, 71.8, 52.4, 58.8 & 6.16E-10, 1.47E-2, .177 \\ \hline
    \end{tabular}
\end{table}

@@

For the multicore tests, averaged speed results appear in \ref{fig:multi_speed}. For this series,
rather than varying the number of nonzero Jacobian matrix elements the number of cores used to
solve the problem is varied, while the number of nonzero Jacobian matrix elements remained constant.
Complex-step again was the slowest method, followed by central-difference, automatic-differentiation
and then forward-difference, taking the title of fastest method for this group of tests.
\ref{fig:multi_speed} indicates that when looking at average in-iteration speed, each of the
methods scale linearly with the number of cores used for the given conditions of this group of
tests. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale
well when additional processors are used. An explanation as to why automatic-differentiation and
forward-differences swapped places in speed ranking relative to the single core results was not
determined.

@@

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in \ref{tcs}. The averaged
results of these accuracy measurements for the single core tests can be seen in
\ref{fig:serial_accuracy}. Averaged Frobenius norm of the element-wise difference between a given
Jacobian and the Jacobian produced by the automatic-differentiation method is plotted against number
of nonzero Jacobian matrix elements. Complex-step was revealed as the most accurate of the methods
besides automatic-differentiation, complex-step resulting in several orders of magnitude less
difference that resulted from using the central-difference method and the most inaccurate
forward-difference method. This accuracy ranking order was mantained for each test in the series. It
should be pointed out compared to the L2 norms of any of the Jacobian matrices by themselves, the
magnitude of these differences for any of the methods is relatively small. These values appear in
the experimental data which is available on the corresponding author's website. The effect of these
accuracy differences on convergence rate directly was not determined.

@@

\subsection{Thoughts on Results} Based on the experimental results, it appears that for users of
\emph{Peridigm} there is a case to be made for using automatic-differentiation for reasons of
performance and accuracy. Although, the greater speed of finite-difference for certain multicore
tests done in this study may give some users pause. While the results showed that complex-step
produced more accurate tangent-stiffness matrices than central-difference and forward-difference 
under the parameters of the tests, it was not determined whether or not this is a clear advantage of complex-step over
finite-difference in terms of accuracy of final predicted node positions and speed of convergence. For other
simulation software it is obviously necessary to collect the same types measurements as decribed in
this paper in order to make a design decision. For new developments it is recommended that combined
accuracy and speed of iteration mesurments be conduted along with entirely separate convergence
studies. For reasons explained in the next subsection, a convergence study in \emph{Peridigm} performed using
the implementations of the methods used for this paper would be moot.

@@

After the bulk of the study had been conducted, gained familiarity with \emph{Peridigm} revealed
that the implementations of forward-difference, central-difference and by virtue of reusing code,
complex-step, in \emph{Peridigm} were not optimal due to inefficiencies in the implementations of
the Linear Peridynamic Solid (LPS) model. Each of these implementations included a large amount of
unnecessary re-calculation of constant values.  The approach used does minimize temporary memory
use, and allow other objects in \emph{Peridigm} to make calls directly to methods in the material
objects, but is wastefull. Unpublished preliminary results show that when calculations for the
constant quantity 'zeta' in the LPS implementation are re-used and only perturbed and previously
pertubed functions of independent variables are recomputed throughout the course of a Jacobian
evaluation, complex-step performs at the speed of unmodified forward-difference. Because of this it
is speculated that forward-difference and central-difference could be modified in nearly the same
way to perform more quickly than AD. In fact, because central-difference involves double the number
of model evaluations than that of complex-step, central-difference stands to gain proportionally
more efficiency. Although no formal academic paper will be likely written based on this, a brief
report and an enhanced performance implementation of finite-difference for peridynamic LPS Jacbian
calculation  in \emph{Peridigm} will eventually be made available on the corresponding author's
website and is intended to be submitted for official inclusion in future releases of \emph{Peridigm}.

\subsection{Final Thoughts on Complex-Step} Complex-step was shown to be highly accurate in
calculating tangent-stiffness matrices by the standards of this study, but as it is implemented in
\emph{Peridigm} may not be any more accurate than AD, and it is almost certainly slower for a naive
implementation. When looking to the future of \emph{Peridigm}, Complex-step holds the advantage of
relying on byte-copiable datatypes, while AD requires a complicated serialization and
deserialization in order to function as byte-copiable. This is fine for copying in between MPI
processes, but deserialization on an accelerator or GPU is very difficult, as this would require
compiling a version of the AD library being used that were compatible with the accelerator or GPU
being used. What this means is that if \emph{Peridigm} or any simulation software were to be
reformulated to take advantage of accelerators or GPUs, attractive given their performance and power
efficiency advantages over traditional CPUs, the flexibilty of complex-step owed to its simple
implementation may make it more viable fo rthat application.
